Simple repression draft

### 2020-05-25 doc/appendix_01_bayesian_inference.tex
update minor things such as figure naming

With this, we are ready to write our full generative model as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2)
\\
\ln b \sim \text{Normal}(0.5, 1)
\\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
We carried out MCMC sampling on the posterior of this model, starting
with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
In constrast to the active/inactive constitutive model considered
in~\cite{Razo-Mejia2020} (nonequilibrium model four
in~\ref{fig:constit_cartoons}), this model is well-identified
with both parameters determined to a fractional uncertainty of 5-10\%.
The strong correlation reflects the fact that their product sets
the mean of the mRNA distribution, which is tightly constrained
by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
along a set of values with a similar product.

>>>>

With this, we are ready to write our full generative model as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2)
\\
\ln b \sim \text{Normal}(0.5, 1)
\\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
We carried out MCMC sampling on the posterior of this model, starting
with the constitutive \textit{lacUV5} dataset from~\cite{Jones2014}.
The resulting MCMC samples are shown in~\fig{fig:constit_post_full}(A).
In constrast to the active/inactive constitutive model considered
in~\cite{Razo-Mejia2020} (nonequilibrium model four
in~\ref{fig2:constit_cartoons}), this model is well-identified
with both parameters determined to a fractional uncertainty of 5-10\%.
The strong correlation reflects the fact that their product sets
the mean of the mRNA distribution, which is tightly constrained
by the data, but there is weak ``sloppiness''~\cite{Transtrum2015}
along a set of values with a similar product.

### 2020-05-25 doc/section_02_means.tex
update minor things such as figure naming

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP per cell and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$
is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4 in~\fig{fig:constit_cartoons}), to
FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and
$k^-$, from which their ratio is estimated to be no more than a few $10^{-1}$
and possibly as small as $10^{-2}$.

>>>>

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well justified on
the basis of the number of RNAP per cell and the strength of binding of RNAP to
DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$
is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
factor? A resolution is offered by an independent inference of $\rho$ in the
absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
(looking ahead, this is equivalent to model 4 in~\fig{fig2:constit_cartoons}), to
FISH data from~\cite{Brewster2014}. This provided a determination of $k^+$ and
$k^-$, from which their ratio is estimated to be no more than a few $10^{-1}$
and possibly as small as $10^{-2}$.

### 2020-05-25 doc/section_03_beyond_means.tex
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4
update minor things such as figure naming

% !TEX root = ./busty_transcription.tex
\section{Beyond Means in Gene Expression}

@@

\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
Before we can tackle simple repression, we need an adequate phenomenological
model of constitutive expression. The literature abounds with options from which
we can choose, and we show several potential kinetic models for constitutive
promoters in~\fig{fig:constit_cartoons}. Let us consider the suitability of each
model for our purposes in turn.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
The left column depicts various plausible models for the dynamics of constitutive promoters.
%We also list literature references where models have been used previously. 
In
model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
(2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
model (2), treating transcription initiation as a multi-step process proceeding
through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
analogous to (2) except with the precise nature of active and inactive states
left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
certain limit of model (4) in which transcripts are produced in bursts, and
initiation of bursts is a Poisson process. 
%\mmnote{Haven't found model 5
%formulated quite like this in the literature, usually burstiness is handled like
%model 4, but I should probably search a bit more.} 
The right column shows the
Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data observed
in~\cite{Jones2014} without introducing additional assumptions and model
complexity.}
\label{fig:constit_cartoons}
\end{figure}

\subsubsection{Noise in the Poisson Promoter Model}
The simplest model of constitutive expression that we can imagine
is shown as model 1 in Figure~\ref{fig:constit_cartoons} and
assumes that
transcripts are produced as a Poisson process from a single promoter state. 
This is the picture from Jones et.\ al.~\cite{Jones2014} that was used to
interpret a systematic study of gene expression noise over a series of
promoters designed to have different strengths.   This
model insists that the ``true'' mRNA distribution is Poisson, implying the Fano
factor $\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute
deviations from Fano = 1 to intensity variability in fluorescent spots, gene
copy number variation, and copy number fluctuations of the transcription
machinery, e.g., RNAP itself. In this picture, the master equation makes no
appearance, and all the corrections to Poisson behavior are derived as additive
corrections to the Fano factor. For disproving the ``universal noise curve''
from So et al~\cite{So2011}, this picture was excellent. It is appealing in its
simplicity, with only two parameters, the initiation rate $r$ and degradation
rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
other words, the model is not excessively complex for the data at hand. But for
many interesting questions, for instance in the recent work~\mmnote{cite
Manuel's preprint once posted}, knowledge of means and variances alone is
insufficient, and a description of the full distribution of molecular counts is
necessary. For this we need a (slightly) more complex model than model 1 that
would allow us to incorporate the non-Poissonian features of constitutive
promoters directly into a master equation formulation.

\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
Our second model of constitutive transcription posits an architecture
in which the promoter is either empty or bound by RNAP
~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
the second row of Figure~\ref{fig:constit_cartoons},  transcription initiation results in
a state transition from the bound to
the unbound state, reflecting the microscopic reality that an RNAP that has
begun to elongate a transcript is no longer available at the start site to begin
another.  As shown in the appendix, the Fano factor in this model is given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\end{align}
The problem with this picture is that the Fano factor is
 always $<1$. 
 To make contact with the experimental reality of $\nu>1$,
clearly some corrections will be needed.
While this model adds an appealing element of microscopic
reality, we are forced to reject it as the additional complexity is unable to
capture the phenomenology of interest. Obviously the promoter state does in fact
proceed through cycles of RNAP binding, initiating, and elongating, but it seems
that the super-Poissonian noise in mRNA copy number we want to model must be
governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.} How might we remedy the deficits of model 2? It is
known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
multitude of distinct steps occur sequentially before RNAP finally escapes into
the elongation phase. Perhaps adding some of this mechanistic detail as shown
in Figure~\ref{fig:constit_cartoons} might
rescue the previous model   The next simplest refinement of that model could consider open
complex formation and promoter escape as separate steps rather than as a single
effective step. In other words, we construct model 3 by adding a single extra
state to model 2, and we will label the two RNAP-bound states as the closed and
open complexes, despite the true biochemical details certainly being more
complex.  For example, earlier work extended this model by adding an additional
repressor bound state and did not explicitly consider the limit with no repressor
that we analyze here~\cite{Mitarai2015}.  Again, our goal here
is not a complete accounting of all the relevant biochemical detail; this is an
exploratory search for the important features that a model needs to include
to square with the known experimental reality of constitutive expression.

>>>>

% !TEX root = ./busty_transcription.tex
\section{Beyond Means in Gene Expression}
\label{sec:beyond_means}

@@

\subsection{A one-state promoter with bursting is the ``best'' model of constitutive promoters}
Before we can tackle simple repression, we need an adequate phenomenological
model of constitutive expression. The literature abounds with options from which
we can choose, and we show several potential kinetic models for constitutive
promoters in~\fig{fig2:constit_cartoons}. Let us consider the suitability of each
model for our purposes in turn.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
The left column depicts various plausible models for the dynamics of constitutive promoters.
%We also list literature references where models have been used previously. 
In
model (1), transcripts are produced in a Poisson process~\cite{Jones2014}. Model
(2) features explicit modeling of RNAP binding/unbinding
kinetics~\cite{Phillips2015a}. Model (3) is a more detailed generalization of
model (2), treating transcription initiation as a multi-step process proceeding
through closed and open complexes~\cite{Mitarai2015}. Model (4) is somewhat
analogous to (2) except with the precise nature of active and inactive states
left ambiguous~\cite{Razo-Mejia2020}. Finally, model (5) can be viewed as a
certain limit of model (4) in which transcripts are produced in bursts, and
initiation of bursts is a Poisson process. 
%\mmnote{Haven't found model 5
%formulated quite like this in the literature, usually burstiness is handled like
%model 4, but I should probably search a bit more.} 
The right column shows the
Fano factor for each model. Note especially the crucial diagnostic: (2) and (3)
have $\nu$ strictly below 1, while only for (4) and (5) can $\nu$ exceed 1.
Models with Fano factors $\le 1$ cannot produce the FISH data observed
in~\cite{Jones2014} without introducing additional assumptions and model
complexity.}
\label{fig2:constit_cartoons}
\end{figure}

\subsubsection{Noise in the Poisson Promoter Model}
The simplest model of constitutive expression that we can imagine
is shown as model 1 in Figure~\ref{fig2:constit_cartoons} and
assumes that
transcripts are produced as a Poisson process from a single promoter state. 
This is the picture from Jones et.\ al.~\cite{Jones2014} that was used to
interpret a systematic study of gene expression noise over a series of
promoters designed to have different strengths.   This
model insists that the ``true'' mRNA distribution is Poisson, implying the Fano
factor $\nu$ must be 1. In~\cite{Jones2014}, the authors carefully attribute
deviations from Fano = 1 to intensity variability in fluorescent spots, gene
copy number variation, and copy number fluctuations of the transcription
machinery, e.g., RNAP itself. In this picture, the master equation makes no
appearance, and all the corrections to Poisson behavior are derived as additive
corrections to the Fano factor. For disproving the ``universal noise curve''
from So et al~\cite{So2011}, this picture was excellent. It is appealing in its
simplicity, with only two parameters, the initiation rate $r$ and degradation
rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
other words, the model is not excessively complex for the data at hand. But for
many interesting questions, for instance in the recent work~\mmnote{cite
Manuel's preprint once posted}, knowledge of means and variances alone is
insufficient, and a description of the full distribution of molecular counts is
necessary. For this we need a (slightly) more complex model than model 1 that
would allow us to incorporate the non-Poissonian features of constitutive
promoters directly into a master equation formulation.

\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}
Our second model of constitutive transcription posits an architecture
in which the promoter is either empty or bound by RNAP
~\cite{Phillips2015a, Phillips2019}. Here,  as shown in 
the second row of Figure~\ref{fig2:constit_cartoons},  transcription initiation results in
a state transition from the bound to
the unbound state, reflecting the microscopic reality that an RNAP that has
begun to elongate a transcript is no longer available at the start site to begin
another.  As shown in the appendix, the Fano factor in this model is given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\end{align}
The problem with this picture is that the Fano factor is
 always $<1$. 
 To make contact with the experimental reality of $\nu>1$,
clearly some corrections will be needed.
While this model adds an appealing element of microscopic
reality, we are forced to reject it as the additional complexity is unable to
capture the phenomenology of interest. Obviously the promoter state does in fact
proceed through cycles of RNAP binding, initiating, and elongating, but it seems
that the super-Poissonian noise in mRNA copy number we want to model must be
governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.} How might we remedy the deficits of model 2? It is
known~\cite{DeHaseth1998} that once RNAP initially binds the promoter region, a
multitude of distinct steps occur sequentially before RNAP finally escapes into
the elongation phase. Perhaps adding some of this mechanistic detail as shown
in Figure~\ref{fig2:constit_cartoons} might
rescue the previous model   The next simplest refinement of that model could consider open
complex formation and promoter escape as separate steps rather than as a single
effective step. In other words, we construct model 3 by adding a single extra
state to model 2, and we will label the two RNAP-bound states as the closed and
open complexes, despite the true biochemical details certainly being more
complex.  For example, earlier work extended this model by adding an additional
repressor bound state and did not explicitly consider the limit with no repressor
that we analyze here~\cite{Mitarai2015}.  Again, our goal here
is not a complete accounting of all the relevant biochemical detail; this is an
exploratory search for the important features that a model needs to include
to square with the known experimental reality of constitutive expression.

### 2020-05-25 doc/section_04_bayesian_inference.tex
update minor things such as figure naming

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig:constit_cartoons}, a one-state bursty model of
constitutive promoters, achieves a Goldilocks level of complexity. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on
single-cell mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig:constit_cartoons}. these models have Fano factors $\nu\le 1$
meaning they are underdisperse relative to Poisson. We will also not explicitly
consider model 4 from~\fig{fig:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

@@

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes
the ``right'' model of a constitutive promoter is the bursty
picture, model five in Figure~\ref{fig:constit_cartoons}.
Therefore our starting point here is the analogous model with
repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
For a given repressor binding site and copy number,
this model has four rate parameters to be inferred:
the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
the initiation rate of bursts, $k_i$, and the mean burst size $b$
(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).

>>>>

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}, a one-state bursty model of
constitutive promoters, achieves a Goldilocks level of complexity. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on
single-cell mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig2:constit_cartoons}. these models have Fano factors $\nu\le 1$
meaning they are underdisperse relative to Poisson. We will also not explicitly
consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

@@

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes
the ``right'' model of a constitutive promoter is the bursty
picture, model five in Figure~\ref{fig2:constit_cartoons}.
Therefore our starting point here is the analogous model with
repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
For a given repressor binding site and copy number,
this model has four rate parameters to be inferred:
the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
the initiation rate of bursts, $k_i$, and the mean burst size $b$
(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).
